Quantifying stochastic noise in cultured\\ circadian reporter cells

### 2015-03-03 main_text.tex
second pass to frank

An detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014}, and has aided the search for small-molecule therapeutics to boost clock amplitude \cite{Chen2013}.

Bioluminescence rhythms at the cell-culture or tissue level are the result of the collective behavior of thousands of cells.
Transcription at the single-cell level is strongly affected by the low molecular counts of the mRNA and protein species involved. 
As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length \cite{Welsh2004}. 
Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues \cite{An2013}.
Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of fibroblast cells and subsequent image processing \cite{Leise2012}. 
As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.

It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
Additionally, it has been shown that in cell cultures which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level \cite{Nagoshi2004, Welsh2004}. 
In this study, we demonstrate single-cell stochasticity can be reliably inferred from the damping rate of population-level bioluminescence recordings if the following conditions are met, which have been shown experimentally for many types of cultured cells \cite{Nagoshi2004, Welsh2004, Guenthner2014, Noguchi2013}.
We assume oscillations in individual cells are
\begin{itemize}
  \item independent (no significant cell-to-cell coupling),
  \item sustained (do not damp on a single-cell basis),
  \item and homogeneous, with no differences in expected period length.
\end{itemize}
We demonstrate the validity and usefulness of such an approach on several types of circadian data.
First, we show using cell-autonomous fibroblast data that stochastic noise is directly related to population-level damping.
Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to single-cell stochasticity is a likely mechanism.
Finally, we calculate the genome-wide effects of siRNA knockdown on single-cell noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
Using this additional information, we show that circadian rhythms have likely evolved to an optimal point of high amplitude and low stochastic noise.
Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on single-cell noise to be quantified in a high-throughput manner.

@@

In this study, we have described a method by which stochastic noise can be estimated from population-level bioluminescence recordings of cultured cells.
Our method assumes independent and sustained oscillators at the single-cell level.
While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured circadian reporters.
Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.

>>>>

A detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014}, and has aided the search for small-molecule therapeutics to boost clock amplitude \cite{Chen2013}.

Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells.
Transcription at the single-cell level is strongly affected by the low molecular counts of the mRNA and protein species involved. 
As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length \cite{Welsh2004}. 
It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
Additionally, it has been shown that in cell cultures which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.

Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues \cite{An2013}.
Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of individual cells and subsequent image processing \cite{Leise2012}. 
As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.

In this study, we demonstrate that changes to single-cell stochasticity can be reliably inferred from the changes to the damping rate of population-level bioluminescence recordings of cultured circadian reporters.
Our method assumes the following conditions are met, which have been shown to hold experimentally for many types of cultured cells \cite{Nagoshi2004, Welsh2004, Guenthner2014, Noguchi2013}:
oscillations in individual cells are
\begin{itemize}
  \item independent (no significant cell-to-cell coupling),
  \item sustained (do not damp on a single-cell basis),
  \item and homogeneous, with no differences in expected period length between cells.
\end{itemize}
We demonstrate the validity and usefulness of such an approach on several types of circadian data.
First, we show using single-cell fibroblast data that stochastic noise is directly related to population-level damping.
Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to single-cell stochasticity is a likely mechanism.
Finally, we calculate the genome-wide effects of siRNA knockdown on single-cell noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
Using this additional information, we show that circadian rhythms have likely evolved to an optimal point of high amplitude and low stochastic noise.
Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on single-cell noise to be quantified in a high-throughput manner.

@@

In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
Our method assumes independent and sustained oscillators at the single-cell level.
While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured fibroblast cells.
Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.
